CrypTool Script \\ Mathematics and Cryptography

### 2002-01-14 script-gb/script-gb.tex
no message

%\pdfinfo{ /Title (CrypTool Tutorial)
%/Creator (TeX)
%/Producer (pdfTeX 0.14a)
%/Author (Deutsch Bank AG)
%/CreationDate (D:20000920201000)
%}

@@

\author{
(c) The authors, 1998-2001 \\
Frankfurt am Main
}

@@

\parskip 4pt
\vskip + 30 pt
{
In this CrypTool tutorial you will find predominantly mathematically oriented
information for using encryption procedures. The main chapters have been written
by various authors and are therefore independent from one another. At the end of
most chapters you will wind literature and URLs.

You will receive information about the principles of symmetrical and
asymmetrical encryption. A large section of the tutorial is dedicated to the
fascinating topic of prime numbers. You will also obtain an overview of the
mathematical ideas behind modern cryptography.

@@

Whereas the program teaches you how to use cryptography in practice, the
tutorial provides those interested in the subject with a deeper understanding of
the mathematical algorithms used.

The authors Bernhard Esslinger, Bartol Filipovic, Henrik Koy and Roger Oyono
would like to take this opportunity to thank their colleagues in the company and
at the universities of Frankfurt, Gie\"sen und Karlsruhe. They particularly thank
Dr. Peer Wichmann from the Karlsruhe computer science research centre
(Forschungszentrum Informatik, FZI) for his down-to-earth support.

\enlargethispage{0.5cm}
As at CrypTool, the quality of the tutorial is enhanced through your suggestions
and ideas for improvement. We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com} or
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
{
\vskip + 7 pt \noindent
The contact persons for this free tool are listed in the readme file for
CrypTool.}
}

@@

This tutorial is delivered together with CrypTool.

@@

For the conventional encryption procedures, automatic analyses are available
that enable you to determine the key with knowledge of the encrypted document
and, possibly, further information (non-encrypted document or language of the
document).\par \vskip + 3pt

@@

A further aim was to enable users to understand the cryptographic procedures
implemented in the Deutsche Bank. In this way, using CrypTool as a reliable
reference implementation of the various encryption procedures, you can test the
encryption implemented in other programs. \par \vskip + 3pt

@@

The purpose of encryption \index{Encryption} is to change data in such a way
that only an authorised recipient is able to reconstruct the plaintext. This has
the advantage that you can transmit encrypted data openly and nevertheless need
not fear a perpetrator reading the data without authorisation. Authorised
recipients possess a piece of secret information --- called the key --- which
allows them to decrypt the data while it remains hidden from everyone else.

One encryption procedure has been proved to be secure --- the {\em One-Time-Pad}
\index{One­Time­Pad}. However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be just as long
as the message to be protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.

For all other procedures there is a (theoretical) possibility of breaking them.
If the procedures are good, however, the time taken to break them is so long
that it is practically impossible to do and these procedures can therefore be
considered (practically) secure.

@@

For {\em symmetric} encryption \index{Encryption!symmetric} the sender and
recipient must possess a common (secret) key which they have exchanged before
actually starting to communicate. The sender uses this key to encrypt the
message and the recipient uses it to decrypt it.

The advantages of symmetric algorithms are the high speed with which data can be
encrypted and decrypted. One disadvantage is the need for key management. In
order to communicate with one another confidentially, sender and recipient must
have exchanged a key using a secure channel before actually starting to
communicate. Spontaneous communication between individuals who have never met
therefore seems virtually impossible. If everyone wants to communicate with
everyone else spontaneously at any time in a network of $ n $ subscribers, each
subscriber must have previously exchanged a key with each of the other $n-­ 1$
subscribers. A total of $n(n - 1)/2$ keys must therefore be exchanged.

The most well-known symmetric encryption procedure is the \index{DES} DES--
algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing all possible keys until the right one is found ({\em brute--force--
attack})\index{Attack!brute-force}. Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.

@@

In the case of {\em asymmetric} encryption \index{Encryption!asymmetric} each
subscriber has a personal pair of keys consisting of a {\em secret}
\index{Key!secret} key and a {\em public} key \index{Key!public}. The public
key, as its name implies, is made public, e.g. in a key directory on the
Internet.

If Alice wants to communicate with Bob, then she finds Bob's public key in the
directory and uses it to encrypt her message to him. She then sends this
ciphertext to Bob, who is then able to decrypt it again using his secret key. As
only Bob knows his secret key, only he can decrypt messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.

Such a procedure can be demonstrated using a series of thief-proof letter boxes.
If I have composed a message, I then look for the letter box of the recipient
and post the letter through it. After that, I can no longer read or change the
message myself, because only the legitimate recipient possesses the key for the
letter box.

The advantage of asymmetric procedures is the easy \index{Key
management} key management. Let's look again at a network with $n$
subscribers. In order to ensure that each subscriber can establish
an encrypted connection to each other subscriber, each subscriber
must possess a pair of keys. We therefore need $2n$ keys or $n$
pairs of keys. Furthermore, no secure channel is needed before
messages are transmitted, because all the information required in
order to communicate confidentially can be transmitted openly. In
this case, you simply have to pay attention to the accuracy
(integrity and authenticity) \index{Authenticity} of the public
key. Disadvantage: Pure asymmetric procedures take a lot longer to
perform than symmetric ones.

The most well-known asymmetric procedure is the \index{RSA} RSA -- algorithm,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA-
algorithm was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signatures!digital}Signature
Algorithm).

\subsubsection{Hybrid procedures}
\index{Hybrid procedures}
In order to benefit from the advantages of symmetric and asymmetric techniques
together, hybrid procedures are usually used (for encryption) in practice.

@@

\newpage
\section{Prime numbers}\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Update Nov. 2000, December 2001)

\paragraph{What are prime numbers?}
\index{Prime numbers} \index{Numbers!prime number}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

The {\em natural} \index{Numbers} numbers thus comprise $1, 2, 3, 4,$ ...
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

@@

The sequence of prime numbers starts with $$ 2,{} 3,{} 5,{} 7, {} 11, {} 13, {}
17, {} 19, {} 23, {} 29, {} 31, {} 37, {} 41, {} 43, {} 47, {} 53, {} 59, {} 61,
{} 67, {} 71, {} 73, {} 79, {} 83, {} 89, {} 97, ... $$ The first 100 numbers
include precisely 25 prime numbers. After this, the percentage of primes
constantly decreases. Prime numbers can be factorised in a unique {\em trivial}
way: $$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1013 = 1 \cdot 1.013,  \quad
1.296.409 = 1 \cdot 1.296.409.$$
All numbers that have $2$ or more factors are called \index{Numbers!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59 $$

@@

In other words: each natural number other than $1$ can be written as a product
of prime numbers in precisely one way, if we ignore the order of the factors.
The factors are therefore unique (the {\em expression as a product of factors}
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot
5^1 $$
And this --- other than changing the order of the factors --- is the only way in
which the number $60$ can be factorised. If you allow numbers other than primes
as factors, there are several ways of factorising integers and the uniqueness \hypertarget{uniqueness}{} is
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = ... $$

The following section is aimed more at those familiar with mathematical logic:
The 1st fundamental theorem only appears to be obvious \label{remFundTheoOfArithm}. We can construct
numerous other sets of numbers (i.e. other than positive whole numbers greater
than 1), for which numbers in the set cannot be expressed uniquely as a product
of the prime numbers of the set: In the set $M = \{1, 5, 10, 15, 20, ...\}$
there is no equivalent to the fundamental theorem under multiplication. The
first five prime numbers of this sequence are $5, 10, 15, 20, 30$ (note: $10$ is
prime, because $5$ is not a factor of $10$ in this set --- the result is not an
element of the given basic set $M$). Because the following applies in $M$: $$
100 = 5 \cdot 20 = 10 \cdot 10 $$ and $5, 10, 20$ are all prime numbers in this
set, the expression as a product of prime factors is not unique here.

@@

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[S. 22]{Blum1999}).

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid knew this in the
third century before Christ. His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\quad Thus there is an {\em infinite} number of primes.
\hyperlink{primhfk}{(Cross-reference: overview under 2.6.1 of the number of
prime numbers in various intervals).}

@@

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ''only'' estimated to be a
$80$-digit number. \hyperlink{grosord}{(See: overview under 2.6.3
of various orders of magnitude / dimensions.)}

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193. 707. 721 \cdot 761, 838, 257, 287 $$
He admitted to having worked 20 years on the factorisation (expression as a
product of factors) of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers have the exponents $$
\begin{array}{c}
2, {} 3, {} 5, {} 7, {} 13, {} 17, {} 19, {} 31, {} 61, {} 89, {} 107, {} 127,
{} 521, {} 607, {} 1.279, {} 2.203, {} 2.281, {} 3.217, {} 4.253, {} 4.423, \\
9.689, {} 9.941, {} 11.213, {} 19.937, {} 21.701, {} 23.207, {} 44.497, {}
86.243, {} 110.503, {} 132.049, {} 216.091,\\
 756.839, {} 859.433, {} 1.257.787, {} 1.398.269, {} 2.976.221, {} 3.021.377, {}
6.972.593, ~13.466.917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known.

@@

These numbers can be found at the following URLs:

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne Marine!M-38}
The 38th Mersenne prime
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 6, 2001} \index{Mersenne Marine!M-39}

The 39th Mersenne prime, called M-39, ($=2^{13.466.917}-1$), was found at December 6, 2001 – more exactly,
he verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron,
was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number (924947738006701322247758 ... 1130073855470256259071) in the Financial Times
would need around 200 pages.

@@

Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search) founded in 1996
already discovered for the 5th time the greatest Mersenne number to be proven being prime.
Now more than 130,000 volunteers, amateurs and experts, are working for the GIMPS project. They connect their
computers into the so called „primenet“, organized by the company entropia, to find such numbers using distributed
computer programs.

@@

The discoverer of M-38 got for discovering the first prime with more than 1 million decimal digits 50.000 USD
from the EFF. The next price from EFF with 100.000 USD is for a proven prime with more than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[''small'' Fermat]
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

{\href{http://www.prothsearch.net/index.html}{\tt
http://www.prothsearch.net/index.html}}.

Cross-reference: the table under 2.6.2 contains the precise values for the $n$th
prime numbers for selected $ n.$

>>>>

%\pdfinfo{ /Title (CrypTool script)
%/Creator (TeX)
%/Producer (pdfTeX 0.14a)
%/Author (Deutsch Bank AG)
%/CreationDate (D:20000920201000)
%}

@@

\author{
(c) The authors, 1998-2002 \\
Frankfurt am Main
}

@@

\parskip 4pt
\vskip + 30 pt
{
In this CrypTool script you will find predominantly mathematically oriented
information for using encryption procedures. The main chapters have been written
by {\em various authors} and are therefore independent from one another. At the end of
most chapters you will find literature and URLs.

You will receive information about the principles of symmetrical and
asymmetrical encryption. A large section of the script is dedicated to the
fascinating topic of prime numbers. With the use of numerous examples, the elementary number
theory and modular arithmetic have been introduced and have been applyed in an exemplary 
manner in the RSA procedure.  After this, one receives insight into the mathematical ideas
behind modern cryptography.

%You will also obtain an overview of the
%mathematical ideas behind modern cryptography.

@@

Whereas the program CrypTool teaches you how to use cryptography in practice, the
script provides those interested in the subject with a deeper understanding of
the mathematical algorithms used -- trying to do it didactically as good as possible.

The {\em authors} Bernhard Esslinger, Bartol Filipovic, Henrik Koy and Roger Oyono
would like to take this opportunity to thank their colleagues in the company and
at the universities of Frankfurt, Gie{\ss}en und Karlsruhe. They particularly thank
Dr. Peer Wichmann from the Karlsruhe computer science research centre
(Forschungszentrum Informatik, FZI) for his down-to-earth support.

\enlargethispage{0.5cm}
As at CrypTool, the quality of the script is enhanced through your suggestions
and ideas for improvement. We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ or~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
{
\vskip + 7 pt \noindent
The contact persons for this free tool are listed in the readme file for
CrypTool.}
}

@@

This script is delivered together with CrypTool.

@@

For the conventional encryption procedures, automatic analyses are available in CrypTool
that enable you to determine the key with knowledge of the encrypted document
and, possibly, further information (non-encrypted document or language of the
document).\par \vskip + 3pt

@@

A further aim was to enable users to understand the cryptographic procedures
implemented in the Deutsche Bank. In this way, using CrypTool as a reliable
reference implementation of the various encryption procedures (because of using the industry-proven SECUDE Library),
you can test the encryption implemented in other programs. \par \vskip + 3pt

Because the articles in this script are largely self-contained, this script can also
be read independantly of CrypTool.

The {\em authors} tried to describe cryptography for a broad audience --
without being mathematically incorrect. We believe, that this didactical pretention
is the way to promote the awareness for IT security and the readiness to use standardised modern cryptography.

@@

The purpose of encryption \index{Encryption} is to change data in such a way
that only an authorised recipient is able to reconstruct the plaintext. This has
the advantage that you can transmit encrypted data openly and nevertheless need
not fear a perpetrator reading the data without authorisation. Authorised
recipients possess a piece of secret information --- called the key --- which
allows them to decrypt the data while it remains hidden from everyone else.\par \vskip + 3pt

One encryption procedure has been proved to be secure --- the {\em One-Time-Pad}.
\index{One­Time­Pad} However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be just as long
as the message to be protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.\par \vskip + 3pt

For all other procedures there is a (theoretical) possibility of breaking them.
If the procedures are good, however, the time taken to break them is so long
that it is practically impossible to do and these procedures can therefore be
considered (practically) secure.\par \vskip + 3pt

@@

For {\em symmetric} encryption \index{Encryption!symmetric} the sender and
recipient must possess a common (secret) key which they have exchanged before
actually starting to communicate. The sender uses this key to encrypt the
message and the recipient uses it to decrypt it.\par \vskip + 3pt

The advantages of symmetric algorithms are the high speed with which data can be
encrypted and decrypted. One disadvantage is the need for key management. In
order to communicate with one another confidentially, sender and recipient must
have exchanged a key using a secure channel before actually starting to
communicate. Spontaneous communication between individuals who have never met
therefore seems virtually impossible. If everyone wants to communicate with
everyone else spontaneously at any time in a network of $ n $ subscribers, each
subscriber must have previously exchanged a key with each of the other $n-­ 1$
subscribers. A total of $n(n - 1)/2$ keys must therefore be exchanged.\par \vskip + 3pt

The most well-known symmetric encryption procedure is the \index{DES} DES--
algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing all possible keys until the right one is found ({\em brute--force--attack}).
\index{Attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.\par \vskip + 3pt

@@

In the case of {\em asymmetric} encryption \index{Encryption!asymmetric} each
subscriber has a personal pair of keys consisting of a {\em secret}
\index{Key!secret} key and a {\em public} key\index{Key!public}. The public
key, as its name implies, is made public, e.g. in a key directory on the
Internet.\par \vskip + 3pt

If Alice wants to communicate with Bob, then she finds Bob's public key in the
directory and uses it to encrypt her message to him. She then sends this
ciphertext to Bob, who is then able to decrypt it again using his secret key. As
only Bob knows his secret key, only he can decrypt messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

Such a procedure can be demonstrated using a series of thief-proof letter boxes.
If I have composed a message, I then look for the letter box of the recipient
and post the letter through it. After that, I can no longer read or change the
message myself, because only the legitimate recipient possesses the key for the
letter box.\par \vskip + 3pt

The advantage of asymmetric procedures is the easy \index{Key
management} key management. Let's look again at a network with $n$
subscribers. In order to ensure that each subscriber can establish
an encrypted connection to each other subscriber, each subscriber
must possess a pair of keys. We therefore need $2n$ keys or $n$
pairs of keys. Furthermore, no secure channel is needed before
messages are transmitted, because all the information required in
order to communicate confidentially can be transmitted openly. In
this case, you simply have to pay attention to the accuracy
(integrity and authenticity) \index{Authenticity} of the public
key. Disadvantage: Pure asymmetric procedures take a lot longer to
perform than symmetric ones.\par \vskip + 3pt

The most well-known asymmetric procedure is the \index{RSA} RSA algorithm,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signatures!digital}Signature
Algorithm).

\newpage
\subsubsection{Hybrid procedures}
\index{Hybrid procedures}
In order to benefit from the advantages of symmetric and asymmetric techniques
together, hybrid procedures are usually used (for encryption) in practice.\par \vskip + 3pt

@@

\newpage
\section{Prime numbers}\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, December 2001)

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime numbers} \index{Numbers!prime number}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

The {\em natural} \index{Numbers} numbers $\mathbb{N}=\{1, 2, 3, 4,\cdots \}$ thus comprise 
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this, the percentage of primes
constantly decreases. Prime numbers can be factorised in a unique {\em trivial}
way: $$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Numbers!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

In other words: each natural number other than $1$ can be written as a product
of prime numbers in precisely one way, if we ignore the order of the factors.
The factors are therefore unique (the {\em expression as a product of factors}
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot
5^1. $$
And this --- other than changing the order of the factors --- is the only way in
which the number $60$ can be factorised. If you allow numbers other than primes
as factors, there are several ways of factorising integers and the uniqueness \hypertarget{uniqueness}{} is
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = \cdots . $$

The following section is aimed more at those familiar with mathematical logic:
The 1st fundamental theorem only appears to be obvious \label{remFundTheoOfArithm}. We can construct
numerous other sets of numbers (i.e. other than positive whole numbers greater
than 1), for which numbers in the set cannot be expressed uniquely as a product
of the prime numbers of the set: In the set $M = \{1, 5, 10, 15, 20, \cdots\}$
there is no equivalent to the fundamental theorem under multiplication. The
first five prime numbers of this sequence are $5, 10, 15, 20, 30$ (note: $10$ is
prime, because $5$ is not a factor of $10$ in this set --- the result is not an
element of the given basic set $M$). Because the following applies in $M$: $$
100 = 5 \cdot 20 = 10 \cdot 10 $$ and $5, 10, 20$ are all prime numbers in this
set, the expression as a product of prime factors is not unique here.

@@

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[p. 22]{Blum1999}).

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{
Euclid, a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ''The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\footnote{
The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. he theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

Thus there is an {\em infinite} number of primes
\hyperlink{primhfk}{(Cross-reference: overview under 2.6.1 of the number of
prime numbers in various intervals)}.\par \vskip + 3pt

Here we make mention of yet another fact which is initially somewhat surprising. 
Namely, in the prime numbers sequence $p_1, p_2, \cdots,$ gaps between prime numbers can have
an individually determined length $n$. It is undeniable that under the $n$
succession of natural numbers
$$(n+1)!+2,\cdots, (n+1)!+(n+1),
$$
none of them is a prime number since in order, the numbers $2,3,\cdots,(n+1)$  
are comprised respectively as real divisors. 
Nevertheless $n!$ means the product of the first $n$ natural numbers therefore 
$ n!= n*(n-1)*\cdots *2*1.$

@@

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ''only'' estimated to be a
$80$-digit number \hyperlink{grosord}{(See: overview under 2.6.3
of various orders of magnitude / dimensions)}.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation (expression as a
product of factors) of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers have the exponents
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253; ~ 4,423; \\
9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; ~ 216,091;\\
 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; ~13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known.

@@

\newpage
These numbers can be found at the following URLs:

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne Marine!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne Marine!M-39}

The 39th Mersenne prime, called M-39, $$2^{13,466,917}-1,$$ was found at December 6, 2001 – more exactly,
the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron,
was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number $(924947738006701322247758 \cdots 1130073855470256259071)$ in the Financial Times
would need around 200 pages.

@@

Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search) founded in 1996
already discovered for the 5th time the greatest Mersenne number to be proven being prime.

Now more than 130,000 volunteers, amateurs and experts, are working for the GIMPS project. They connect their
computers into the so called ''primenet'', organized by the company entropia, to find such numbers using distributed
computer programs.

@@

The discoverer of M-38 got for discovering the first prime with more than 1 million decimal digits 50,000 USD
from the EFF. The next price from EFF with 100,000 USD is for a proven prime with more than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[''small'' Fermat]
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

{\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
 {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.

Cross-reference:  \hyperlink{ntePrimzahl}{the table under 2.6.2} contains the precise values for the $n$th
prime numbers for selected $ n.$
